Hazelnut: A Bidirectionally Typed \\ Structure Editor Calculus

### 2016-06-22 hazelnut-popl17.tex
updated abstract

\def\OPTIONConf{0}%
%\documentclass[12pt]{article}
\documentclass[preprint,9pt]{sigplanconf}
% The following \documentclass options may be useful:
%
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% authoryear    To obtain author/year citation style instead of numeric.
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{ stmaryrd }
\usepackage{mathpartir}
\usepackage{stmaryrd}
\usepackage{wasysym}
\usepackage{extarrows}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\usepackage{mathpazo}
\usepackage{colortab}
\usepackage{url}
\usepackage{todonotes}
\sloppy
\def \TirNameStyle #1{\small\rulename{#1}}
\renewcommand{\MathparLineskip}{\lineskiplimit=.3\baselineskip\lineskip=.3\baselineskip plus .2\baselineskip}

@@

\maketitle
\begin{abstract}
Programmers typically construct and manipulate well-typed expressions  only indirectly, as text that must pass through a parser and type\-checker. Not all text survives this journey. In particular, text that arises transiently during the editing process, or when the programmer has made a mistake, is often malformed or ill-typed. 
% Not all text that arises during the programming process survives the journey through a parser and typechecker. In particular, text that arises transiently during the editing process, or when the programmer has made a mistake, is often malformed or ill-typed. 
Contending with (i.e. reasoning about) malformed text, or well-formed but meaningless expressions, is a difficult task, for programmers and tools alike.

%  \emph{Structure editors} have long promised to alleviate these burdens by exposing only edit actions that  cause sensible changes to the program structure.
% Existing designs for structure editors, however, are complex and somewhat \emph{ad hoc}. They also focus primarily on syntactic well-formedness, so programs can still be left semantically meaningless as they are being constructed.

This paper introduces Hazelnut, a minimal \emph{structure editor} where every edit action leaves the expression being edited in both a statically and dynamically meaningful state (with the latter implying that Hazelnut is particularly suitable as a basis for \emph{live programming}.) Uniquely, 1) Hazelnut is defined in a principled type-theoretic style, with a rich metatheory mechanized in Agda; and 2) users need not construct outer forms before they construct inner forms. Instead, Hazelnut's \emph{action semantics} automatically places terms that have a type that is inconsistent with the expected type into a \emph{hole}, which safely defers the type consistency check until the hole is \emph{finished}. Interestingly, this mechanism is related to a mechanism that arises in the semantics of gradual type systems. %Formally, Hazelnut is a bidirectionally typed lambda calculus extended with \emph{holes}, a \emph{focus model} (based on Huet's zipper) and an \emph{action model}.
\end{abstract}

@@

The rules for relative movement within Z-expressions are similar. Movement is type-independent, so we defer to an auxiliary judgement for both the analytic and synthetic judgements:
\begin{subequations}
\begin{equation}
\inferrule{
  \performMove{\zexp}{\aMove{\delta}}{\zexp'}
}{
  \performSyn{\hGamma}{\zexp}{\htau}{\aMove{\delta}}{\zexp'}{\htau}
}
\end{equation}
\begin{equation}
  \inferrule{
  \performMove{\zexp}{\aMove{\delta}}{\zexp'}
}{
  \performAna{\hGamma}{\zexp}{\htau}{\aMove{\delta}}{\zexp'}
}
\end{equation}
\end{subequations}
For concision, we show only the rules for ascription here:
\begin{subequations}
  \begin{equation}
    \label{r:movefirstchild}
  \inferrule{ }{
    \performTyp{
      \zwsel{\hexp : \htau}
    }{
      \aMove{\dChild}
    }{
      \zwsel{\hexp} : \htau
    }
  }
\end{equation}
\begin{equation}
  \label{r:moveparent}
  \inferrule{ }{
    \performTyp{
      \zwsel{\hexp} : \htau
    }{
      \aMove{\dParent}
    }{
      \zwsel{\hexp : \htau}
    }
  }
\end{equation}
\begin{equation}
  \inferrule{ }{
    \performTyp{
      \hexp : \zwsel{\htau}
    }{
      \aMove{\dParent}
    }{
      \zwsel{\hexp : \htau}
    }
  }
\end{equation}
\begin{equation}
  \label{r:movenextsib}
  \inferrule{ }{
    \performTyp{
      \zwsel{\hexp} : \htau
    }{
      \aMove{\dNext}
    }{
      \hexp : \zwsel{\htau}
    }
  }
\end{equation}
\begin{equation}
  \label{r:moveprevsib}
  \inferrule{ }{
    \performTyp{
      \hexp : \zwsel{\htau}
    }{
      \aMove{\dPrev}
    }{
      \zwsel{\hexp} : \htau
    }
  }
\end{equation}
\begin{equation}
\inferrule{
  \performTyp{
    \zexp
  }{
    \aMove{\delta}
  }{
    \zexp'
  }
}{
  \performTyp{
    \zexp : \htau
  }{
    \aMove{\delta}
  }{
    \zexp' : \htau
  }
}
\end{equation}
\begin{equation}
  \inferrule{
    \performTyp{
      \ztau
    }{
      \aMove{\delta}
    }{
      \ztau'
    }
  }{
    \performTyp{
      \hexp : \ztau
    }{
      \aMove{\delta}
    }{
      \hexp : \ztau'
    }
  }
\end{equation}
\end{subequations}
\subsubsection{Deletion} The $\aDel$ action replaces the selected subterm with an empty hole.

@@

>>>>

\def\OPTIONConf{0}%
%\documentclass[12pt]{article}
\documentclass[preprint,9pt]{sigplanconf}
% The following \documentclass options may be useful:
%
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% authoryear    To obtain author/year citation style instead of numeric.
\usepackage[T1]{fontenc}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{ stmaryrd }
\usepackage{mathpartir}
\usepackage{stmaryrd}
\usepackage{wasysym}
\usepackage{extarrows}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\usepackage{mathpazo}
\usepackage{colortab}
\usepackage{url}
\usepackage{todonotes}
\usepackage{listings}
\usepackage{microtype}
\sloppy
\def \TirNameStyle #1{\small\rulename{#1}}
\renewcommand{\MathparLineskip}{\lineskiplimit=.3\baselineskip\lineskip=.3\baselineskip plus .2\baselineskip}

@@

\maketitle
\begin{abstract}
% Programmers typically construct and manipulate well-typed expressions  only indirectly, as text that must pass through a parser and type\-checker. Not all text survives this journey. In particular, text that arises transiently, or when the programmer makes a mistake, is often malformed or ill-typed. 
% Not all text that arises during the programming process survives the journey through a parser and typechecker. In particular, text that arises transiently during the editing process, or when the programmer has made a mistake, is often malformed or ill-typed. 
% Contending with malformed program text or well-formed but meaningless syntax trees is difficult for programmers and their tools alike.

%  \emph{Structure editors} have long promised to alleviate these burdens by exposing only edit actions that  cause sensible changes to the program structure.
% Existing designs for structure editors, however, are complex and somewhat \emph{ad hoc}. They also focus primarily on syntactic well-formedness, so programs can still be left semantically meaningless as they are being constructed.
% Structure editors (e.g. Scratch) eliminate the possibility of syntax errors.

\emph{Structure editors} allow programmers to edit the tree structure of a program directly -- there are no syntax errors. This can have cognitive benefits, particularly for novice and end-user programmers (as evidenced by the popularity of structure editors like Scratch.) It also simplifies matters for tool designers, because they need not contend with malformed program text. %Structure editors for professional programmers are also under active development (e.g. \texttt{mbeddr}.)

This paper defines Hazelnut, a {structure editor} based on a small bidirectionally typed lambda calculus that takes this approach two steps further: Hazelnut's edit actions operate over terms that are not just syntactically well-formed but also statically and dynamically meaningful. 
Na\"ively, this injunction against type errors would force programmers to construct terms in an unnatural ``outside-in'' manner. To avoid this problem, Hazelnut's {action semantics} automatically places terms that have a type that is incompatible with the expected type into a \emph{hole}, which safely defers the type compatibility check until the hole is \emph{finished}. Interestingly, this mechanism is related to a mechanism that arises in gradual type systems.

We intend Hazelnut as a minimal foundational calculus, rather than as an end-user tool itself. To that end, we show how Hazelnut's rich metatheory, which we have mechanized in Agda, guides the definition of an extension to Hazelnut. We also briefly discuss how Hazelnut's semantics lends itself to implementation as a functional reactive web application. Our reference implementation is written using \lstinline{js_of_ocaml}.

%Formally, Hazelnut is a bidirectionally typed lambda calculus extended with \emph{holes}, a \emph{focus model} (based on Huet's zipper) and an \emph{action model}.
\end{abstract}

@@

The rules for relative movement within Z-expressions are similar. Movement is type-independent, so we defer to an auxiliary judgement for both the analytic and synthetic judgements:
\begin{subequations}
\begin{equation}
\inferrule{
  \performMove{\zexp}{\aMove{\delta}}{\zexp'}
}{
  \performSyn{\hGamma}{\zexp}{\htau}{\aMove{\delta}}{\zexp'}{\htau}
}
\end{equation}
\begin{equation}
  \inferrule{
  \performMove{\zexp}{\aMove{\delta}}{\zexp'}
}{
  \performAna{\hGamma}{\zexp}{\htau}{\aMove{\delta}}{\zexp'}
}
\end{equation}
\end{subequations}
For concision, we show only the rules for ascription here:
\begin{subequations}
  \begin{equation}
    \label{r:movefirstchild}
  \inferrule{ }{
    \performTyp{
      \zwsel{\hexp : \htau}
    }{
      \aMove{\dChild}
    }{
      \zwsel{\hexp} : \htau
    }
  }
\end{equation}
\begin{equation}
  \label{r:moveparent}
  \inferrule{ }{
    \performTyp{
      \zwsel{\hexp} : \htau
    }{
      \aMove{\dParent}
    }{
      \zwsel{\hexp : \htau}
    }
  }
\end{equation}
\begin{equation}
  \inferrule{ }{
    \performTyp{
      \hexp : \zwsel{\htau}
    }{
      \aMove{\dParent}
    }{
      \zwsel{\hexp : \htau}
    }
  }
\end{equation}
\begin{equation}
  \label{r:movenextsib}
  \inferrule{ }{
    \performTyp{
      \zwsel{\hexp} : \htau
    }{
      \aMove{\dNext}
    }{
      \hexp : \zwsel{\htau}
    }
  }
\end{equation}
\begin{equation}
  \label{r:moveprevsib}
  \inferrule{ }{
    \performTyp{
      \hexp : \zwsel{\htau}
    }{
      \aMove{\dPrev}
    }{
      \zwsel{\hexp} : \htau
    }
  }
\end{equation}
\begin{equation}
\inferrule{
  \performTyp{
    \zexp
  }{
    \aMove{\delta}
  }{
    \zexp'
  }
}{
  \performTyp{
    \zexp : \htau
  }{
    \aMove{\delta}
  }{
    \zexp' : \htau
  }
}
\end{equation}
\begin{equation}
  \inferrule{
    \performTyp{
      \ztau
    }{
      \aMove{\delta}
    }{
      \ztau'
    }
  }{
    \performTyp{
      \hexp : \ztau
    }{
      \aMove{\delta}
    }{
      \hexp : \ztau'
    }
  }
\end{equation}
and this for the example
\begin{equation}\label{r:movefirstchild-lam}
\inferrule{ }{
  \performMove{
    \zwsel{\hlam{x}{\hexp}}
  }{
    \aMove{\dChild}
  }{
    \hlam{x}{\zwsel{\hexp}}
  }
}
\end{equation}

\begin{equation}\label{r:moveparent-ap2}
  \inferrule{ }{
    \performMove{
      \hap{{\hexp_1}}{\zwsel{\hexp_2}}
    }{
      \aMove{\dParent}
    }{
      \zwsel{\hap{\hexp_1}{\hexp_2}}
    }
  }
\end{equation}

\begin{equation}\label{r:moveparent-hole}
  \inferrule{ }{
    \performMove{
      \hhole{\zwsel{\hexp}}
    }{
      \aMove{\dParent}
    }{
      \zwsel{\hhole{\hexp}}
    }
  }
\end{equation}

\end{subequations}
\subsubsection{Deletion} The $\aDel$ action replaces the selected subterm with an empty hole.

@@

TODO: \url{http://research.nii.ac.jp/~hu/pub/hosc07.pdf}

### 2016-06-22 intro.tex
updated abstract

%Although all programs can be represented as text, most text does not correspond to a syntactically well-formed and semantically well-defined program. 
Programmers and the tools that they use often encounter text that does not correspond to a meaningful program. This may be because the programmer is in the midst of a sequence of 
edit actions that leaves the text temporarily malformed or ill-typed, or because the programmer has made a mistake. It is difficult for tools to help programmers execute a corrective course of action in such situations (e.g. by providing syntax highlighting and semantics-aware code completion.) %because the language definition is necessarily silent about the structure or meaning of such text. %For example, a code editor cannot assist the programmer in these situations.

Some editors have developed \emph{ad hoc} workarounds for this problem, e.g. they might use whitespace to guess where a construct is likely to end, attempt to insert closing delimiters automatically, or recover from a type error by pretending that the type was as expected (if, indeed, an expected type can be determined.\todo{citations?}) These heuristic methods are quite complex and sometimes misleading.

@@

\emph{Structure editors} have made some progress toward addressing this
problem by allowing programmers to edit the tree structure of a program
directly. Each edit action leaves the program being constructed in a
structurally well-formed state.\footnote{In addition to eliminating malformed edit states, it is also generally the case that programs can be written more quickly using a structure editor. However, we will not talk about ``edit costs'' here, because we would like to abstract away from details like whether a keyboard or a mouse (or some other input device) is being used.} We give examples of notable structure editors in Sec. \ref{sec:rw}.

>>>>

%Although all programs can be represented as text, most text does not correspond to a syntactically well-formed and semantically well-defined program. 
Programmers and the tools that they use often encounter text that does not correspond to a meaningful program. This may be because the programmer is in the midst of a sequence of 
edit actions that leaves the text temporarily malformed or ill-typed, or because the programmer has made a mistake\todo{mention YoungSeok's data here?}. It is difficult for tools to help programmers determine and execute a corrective course of action in such situations (e.g. by providing \todo{cite study on benefits of syntax highlighting}syntax highlighting and semantics-aware code completion services\todo{cite something?}.) %because the language definition is silent %For example, a code editor cannot assist the programmer in these situations.

Some editors have developed \emph{ad hoc} workarounds for this problem, e.g. they might attempt to insert closing delimiters automatically, use whitespace to guess where a construct is likely to end, or recover from a type error by pretending that the type was as expected (if, indeed, an expected type can be determined.\todo{citations?}) These heuristic methods are imperfect, complex and sometimes misleading.

@@

\emph{Structure editors} have long promised to address this
problem by allowing programmers to edit the tree structure of a program, or some rich projection thereof, 
directly. Each edit action leaves the program being constructed in a
structurally well-formed state.\footnote{In addition to eliminating malformed edit states, it is also often promised that programs can be written more quickly using a structure editor. However, we will not discuss such ``edit costs'' here, because they depend on particular implementation details, e.g. whether a keyboard or a mouse is in use.} We give examples of notable structure editors in Sec. \ref{sec:rw}.
